---
title: "Heterogeneous Tontine: Convex Combination of Sharing Rules"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
library(readxl)
```

# Data and functions

```{r functions}
# --- 1. Load the mortality table (CPM2014, Table 1) --------------------------
Data_raw <- read_excel(path = 'CPM2014 Mortality Table.xlsx')
Mortality_Table <- data.frame(
  age       = as.numeric(Data_raw$Age),
  qx_male   = as.numeric(Data_raw$Male),
  qx_female = as.numeric(Data_raw$Female)
)
rm(Data_raw)
Mortality_Table <- Mortality_Table[!is.na(Mortality_Table$age), ]

# --- 2. qx lookup -----------------------------------------------------------
get_qx <- function(ages, sexes) {
  index <- match(ages, Mortality_Table$age)
  if (any(is.na(index))) warning("Some ages are outside the table")
  ifelse(sexes == 'M',
         Mortality_Table$qx_male[index],
         Mortality_Table$qx_female[index])
}

# --- 3. Portfolio generator -------------------------------------------------
generate_portfolio <- function(n, age_min = 55, age_max = 90,
                               contrib = 100, contrib_sd = 0,
                               prob_male = 0.5, age_dist = "uniform",
                               age_mean = 70, age_sd = 8, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  sex <- ifelse(runif(n) < prob_male, "M", "F")
  ages_range <- age_min:age_max
  age <- switch(age_dist,
    uniform = sample(ages_range, n, replace = TRUE),
    normal  = round(pmax(age_min, pmin(age_max, rnorm(n, age_mean, age_sd)))),
    peaked  = sample(ages_range, n, replace = TRUE,
                     prob = dnorm(ages_range, age_mean, age_sd)),
    stop("unknown age_dist: ", age_dist)
  )
  a <- if (contrib_sd > 0) round(pmax(1, rnorm(n, contrib, contrib_sd)))
       else rep(contrib, n)
  data.frame(id = 1:n, age = age, sex = sex,
             qx = get_qx(age, sex), contributions = a,
             stringsAsFactors = FALSE)
}

# --- 4. Distribution of S_n by convolution ----------------------------------
convolution2 <- function(d1, d2) {
  values <- outer(d1$value, d2$value, "+")
  probs  <- outer(d1$prob,  d2$prob,  "*")
  merged <- tapply(probs, round(values, 8), sum)
  data.frame(value = as.numeric(names(merged)), prob = as.numeric(merged))
}

pool_distribution <- function(tontine_data, rows = seq_len(nrow(tontine_data))) {
  one_member <- lapply(rows, function(j)
    data.frame(value = c(0, tontine_data$contributions[j]),
               prob  = c(1 - tontine_data$qx[j], tontine_data$qx[j])))
  Reduce(convolution2, one_member)
}

# --- 5. CMRS rule (Denuit-Hieber-Robert, formula 2.2) -----------------------
cmrs_one <- function(pf, i, full = pool_distribution(pf)) {
  loo  <- pool_distribution(pf, setdiff(seq_len(nrow(pf)), i))
  p_loo <- loo$prob[match(round(full$value - pf$contributions[i], 8),
                          round(loo$value, 8))]
  p_loo[is.na(p_loo)] <- 0
  data.frame(id = pf$id[i], s = full$value, prob = full$prob,
             credit = pf$contributions[i] * pf$qx[i] * p_loo / full$prob)
}

# --- 6. Sharing rules ---------------------------------------------------
# Proportional
prop_one <- function(pf, i, full = pool_distribution(pf)) {
  w_i  <- (pf$qx[i] * pf$contributions[i]) / sum(pf$qx * pf$contributions)
  data.frame(id = pf$id[i], s = full$value, prob = full$prob,
             credit = w_i * full$value)
}

# Regression rule (weights by variance sigma_i^2)
regression_one <- function(pf, i, full = pool_distribution(pf)) {
  s <- full$value; p <- full$prob
  ES  <- sum(s * p)
  EXi <- pf$qx[i] * pf$contributions[i]
  sigma_i2  <- pf$contributions[i]^2 * pf$qx[i] * (1 - pf$qx[i])
  sigma_sum <- sum(pf$contributions^2 * pf$qx * (1 - pf$qx))
  data.frame(id = pf$id[i], s = s, prob = p,
             credit = EXi + (sigma_i2 / sigma_sum) * (s - ES))
}

# Pure-endowment rule (weight c_i / survival prob)
endowment_one <- function(pf, i, full = pool_distribution(pf)) {
  s <- full$value; p <- full$prob
  ES  <- sum(s * p)
  EXi <- pf$qx[i] * pf$contributions[i]
  pi_surv <- 1 - pf$qx
  w <- (pf$contributions[i] / pi_surv[i]) / sum(pf$contributions / pi_surv)
  data.frame(id = pf$id[i], s = s, prob = p,
             credit = EXi + w * (s - ES))
}

# --- 7. Convex combination --------------------------------------------------
combo_one <- function(pf, i, delta) {
  cm <- cmrs_one(pf, i)
  pr <- prop_one(pf, i)
  data.frame(id = pf$id[i], s = cm$s, prob = cm$prob,
             credit = delta * cm$credit + (1 - delta) * pr$credit)
}

# --- 8. CRRA utility and EXACT expected utility -----------------------------
utility <- function(w, gamma) {
  if (abs(gamma - 1) < 1e-9) log(w) else (w^(1 - gamma)) / (1 - gamma)
}

expected_utility_exact <- function(pf, i, delta, gamma, beta = 1) {
  a_i <- pf$contributions[i]; q_i <- pf$qx[i]
  cc  <- combo_one(pf, i, delta)
  loo <- pool_distribution(pf, setdiff(seq_len(nrow(pf)), i))
  idx_survive <- match(round(loo$value,       8), round(cc$s, 8))
  idx_die     <- match(round(loo$value + a_i, 8), round(cc$s, 8))
  credit_survive <- ifelse(is.na(idx_survive), 0, cc$credit[idx_survive])
  credit_die     <- ifelse(is.na(idx_die),     0, cc$credit[idx_die])
  U_survive <- sum(utility(a_i + credit_survive, gamma) * loo$prob)
  U_die     <- sum(utility(credit_die,           gamma) * loo$prob)
  (1 - q_i) * U_survive + beta * q_i * U_die
}

# --- 8b.(heavy parts computed once per member) ------
# precompute_member() runs the expensive convolutions ONCE; eu_fast() then
# evaluates expected utility for any (delta, gamma, beta).
# eu_fast() is numerically identical to expected_utility_exact().
precompute_member <- function(pf, i,
    rules_list = list(cmrs = cmrs_one, prop = prop_one,
                      regression = regression_one, endowment = endowment_one)) {
  full <- pool_distribution(pf)
  a_i <- pf$contributions[i]; q_i <- pf$qx[i]
  credit_mat <- sapply(rules_list, function(rule) rule(pf, i, full)$credit)
  loo <- pool_distribution(pf, setdiff(seq_len(nrow(pf)), i))
  list(a_i = a_i, q_i = q_i, s = full$value,
       credit_mat = credit_mat, rule_names = names(rules_list),
       prob = loo$prob,
       idx_s = match(round(loo$value,       8), round(full$value, 8)),
       idx_d = match(round(loo$value + a_i, 8), round(full$value, 8)))
}

# delta is a vector, one weight per rule, summing to 1.
eu_fast <- function(P, delta, gamma, beta = 1) {
  credit <- as.vector(P$credit_mat %*% delta)
  cs <- ifelse(is.na(P$idx_s), 0, credit[P$idx_s])
  cd <- ifelse(is.na(P$idx_d), 0, credit[P$idx_d])
  (1 - P$q_i) * sum(utility(P$a_i + cs, gamma) * P$prob) +
    beta * P$q_i * sum(utility(cd, gamma) * P$prob)
}
```

# The portfolio

```{r portfolio}
pf <- generate_portfolio(n = 100, age_min = 60, age_max = 85,
                         contrib = 50, contrib_sd = 8,
                         prob_male = 0.5, age_dist = 'peaked',
                         age_mean = 70, age_sd = 8, seed = 125)
head(pf)
summary(pf$age)
table(pf$sex)
mean(pf$qx)
```




## Portfolio composition

```{r portfolio-viz, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))

hist(pf$age, breaks = 15, col = "grey80", border = "white",
     main = "Age distribution", xlab = "Age")

barplot(table(pf$sex), col = c("steelblue", "salmon"),
        main = "Sex composition", ylab = "Count")

boxplot(age ~ sex, data = pf, col = c("salmon", "steelblue"),
        main = "Age by sex", xlab = "Sex", ylab = "Age")

hist(pf$qx, breaks = 15, col = "grey80", border = "white",
     main = "Death-probability (qx) distribution", xlab = "qx")

par(mfrow = c(1, 1))
```

# Expected utility vs mixing weight, for several gamma


```{r multi-gamma-panels, fig.width=9, fig.height=5}
# 1. EU vs delta, several gammas (one row of panels)
plot_member <- function(pf, id, gammas = c(1.5, 2.5, 3),
                        rules = c("cmrs", "prop"), beta = 1, step = 0.02) {
  i <- match(id, pf$id); P <- precompute_member(pf, i)
  k1 <- match(rules[1], P$rule_names); k2 <- match(rules[2], P$rule_names)
  deltas <- seq(0, 1, by = step)
  label <- sprintf("id %d: %s, age %d, inv %g", id, pf$sex[i], pf$age[i], pf$contributions[i])
  par(mfrow = c(1, length(gammas)), mar = c(4, 4, 3, 1))
  for (g in gammas) {
    eu <- sapply(deltas, function(d) {
      w <- numeric(length(P$rule_names)); w[k1] <- d; w[k2] <- 1 - d
      eu_fast(P, w, gamma = g, beta = beta) })
    d_star <- deltas[which.max(eu)]
    plot(deltas, eu, type = "l", lwd = 2,
         xlab = bquote(delta ~ "(" * .(rules[1]) * ")"), ylab = "Expected utility",
         main = bquote(.(label) ~ "|" ~ gamma == .(g) ~ "," ~ delta*"*" == .(d_star)))
    abline(v = d_star, col = "red", lty = 2)
  }
  par(mfrow = c(1, 1))
}
plot_member(pf, id = 16, rules=c("cmrs","prop"))
plot_member(pf, id = 32, rules=c("cmrs","prop"))
```

# Optimal mixing weight as a function of risk aversion

For each `gamma`, the value of `delta` maximizing expected utility. This shows whether the optimal blend of CMRS and proportional shifts
with risk aversion, and whether it stays interior or hits a corner.

```{r optimal-delta, fig.width=9, fig.height=5}
# 2. Optimal delta* vs gamma
plot_optimal_delta <- function(pf, id, rules = c("cmrs", "prop"),
                               gammas = seq(0.5, 6, by = 0.5), beta = 1, step = 0.01) {
  i <- match(id, pf$id); P <- precompute_member(pf, i)
  k1 <- match(rules[1], P$rule_names); k2 <- match(rules[2], P$rule_names)
  deltas <- seq(0, 1, by = step)
  label <- sprintf("id %d: %s, age %d, inv %g", id, pf$sex[i], pf$age[i], pf$contributions[i])
  opt <- sapply(gammas, function(g) {
    eu <- sapply(deltas, function(d) {
      w <- numeric(length(P$rule_names)); w[k1] <- d; w[k2] <- 1 - d
      eu_fast(P, w, gamma = g, beta = beta) })
    deltas[which.max(eu)] })
  plot(gammas, opt, type = "b", pch = 19, lwd = 2, ylim = c(0, 1),
       xlab = expression(gamma), ylab = bquote(delta*"*" ~ "(" * .(rules[1]) * ")"),
       main = paste0(label, ": optimal delta vs gamma"))
  abline(h = c(0, 1), lty = 3, col = "grey60")
}
plot_optimal_delta(pf, id = 16, rules=c("cmrs","prop"))
```

# Effect of the bequest weight

The same optimal-`delta` curve for two bequest weights `beta`, to see how a
weaker bequest motive changes the optimal blend.

```{r bequest, fig.width=9, fig.height=5}
# 3. Optimal delta* vs gamma, two bequest weights
plot_bequest <- function(pf, id, rules = c("cmrs", "prop"),
                         gammas = seq(0.5, 6, by = 0.5), betas = c(1, 0.2), step = 0.01) {
  i <- match(id, pf$id); P <- precompute_member(pf, i)
  k1 <- match(rules[1], P$rule_names); k2 <- match(rules[2], P$rule_names)
  deltas <- seq(0, 1, by = step)
  label <- sprintf("id %d: %s, age %d, inv %g", id, pf$sex[i], pf$age[i], pf$contributions[i])
  opt_for_beta <- function(b) sapply(gammas, function(g) {
    eu <- sapply(deltas, function(d) {
      w <- numeric(length(P$rule_names)); w[k1] <- d; w[k2] <- 1 - d
      eu_fast(P, w, gamma = g, beta = b) })
    deltas[which.max(eu)] })
  o1 <- opt_for_beta(betas[1]); o2 <- opt_for_beta(betas[2])
  plot(gammas, o1, type = "b", pch = 19, lwd = 2, ylim = c(0, 1),
       xlab = expression(gamma), ylab = expression(delta*"*"),
       main = paste0(label, ": optimal delta vs gamma, by bequest"))
  lines(gammas, o2, type = "b", pch = 17, lwd = 2, col = "red")
  legend("bottomleft", legend = paste0("beta = ", betas),
         col = c("black", "red"), pch = c(19, 17), lwd = 2, bty = "n")
}
plot_bequest(pf, id = 16, rules=c("cmrs","prop"))
```
