---
title: "Heterogeneous Tontine: Convex Combination of Sharing Rules"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
library(readxl)
```

# Data and functions

```{r functions}
# --- 1. Load the mortality table (CPM2014, Table 1) --------------------------
Data_raw <- read_excel(path = 'CPM2014 Mortality Table.xlsx')
Mortality_Table <- data.frame(
  age       = as.numeric(Data_raw$Age),
  qx_male   = as.numeric(Data_raw$Male),
  qx_female = as.numeric(Data_raw$Female)
)
rm(Data_raw)
Mortality_Table <- Mortality_Table[!is.na(Mortality_Table$age), ]

# --- 2. qx lookup -----------------------------------------------------------
get_qx <- function(ages, sexes) {
  index <- match(ages, Mortality_Table$age)
  if (any(is.na(index))) warning("Some ages are outside the table")
  ifelse(sexes == 'M',
         Mortality_Table$qx_male[index],
         Mortality_Table$qx_female[index])
}

# --- 3. Portfolio generator -------------------------------------------------
generate_portfolio <- function(n, age_min = 55, age_max = 90,
                               contrib = 100, contrib_sd = 0,
                               prob_male = 0.5, age_dist = "uniform",
                               age_mean = 70, age_sd = 8, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  sex <- ifelse(runif(n) < prob_male, "M", "F")
  ages_range <- age_min:age_max
  age <- switch(age_dist,
    uniform = sample(ages_range, n, replace = TRUE),
    normal  = round(pmax(age_min, pmin(age_max, rnorm(n, age_mean, age_sd)))),
    peaked  = sample(ages_range, n, replace = TRUE,
                     prob = dnorm(ages_range, age_mean, age_sd)),
    stop("unknown age_dist: ", age_dist)
  )
  a <- if (contrib_sd > 0) round(pmax(1, rnorm(n, contrib, contrib_sd)))
       else rep(contrib, n)
  data.frame(id = 1:n, age = age, sex = sex,
             qx = get_qx(age, sex), contributions = a,
             stringsAsFactors = FALSE)
}

# --- 4. Distribution of S_n by convolution ----------------------------------
convolution2 <- function(d1, d2) {
  values <- outer(d1$value, d2$value, "+")
  probs  <- outer(d1$prob,  d2$prob,  "*")
  merged <- tapply(probs, round(values, 8), sum)
  data.frame(value = as.numeric(names(merged)), prob = as.numeric(merged))
}

pool_distribution <- function(tontine_data, rows = seq_len(nrow(tontine_data))) {
  one_member <- lapply(rows, function(j)
    data.frame(value = c(0, tontine_data$contributions[j]),
               prob  = c(1 - tontine_data$qx[j], tontine_data$qx[j])))
  Reduce(convolution2, one_member)
}

# --- 5. CMRS rule (Denuit-Hieber-Robert, formula 2.2) -----------------------
cmrs_one <- function(pf, i, full = pool_distribution(pf)) {
  loo  <- pool_distribution(pf, setdiff(seq_len(nrow(pf)), i))
  p_loo <- loo$prob[match(round(full$value - pf$contributions[i], 8),
                          round(loo$value, 8))]
  p_loo[is.na(p_loo)] <- 0
  data.frame(id = pf$id[i], s = full$value, prob = full$prob,
             credit = pf$contributions[i] * pf$qx[i] * p_loo / full$prob)
}

# --- 6. Proportional rule ---------------------------------------------------
prop_one <- function(pf, i) {
  full <- pool_distribution(pf)
  w_i  <- (pf$qx[i] * pf$contributions[i]) / sum(pf$qx * pf$contributions)
  data.frame(id = pf$id[i], s = full$value, prob = full$prob,
             credit = w_i * full$value)
}

# --- 7. Convex combination --------------------------------------------------
combo_one <- function(pf, i, delta) {
  cm <- cmrs_one(pf, i)
  pr <- prop_one(pf, i)
  data.frame(id = pf$id[i], s = cm$s, prob = cm$prob,
             credit = delta * cm$credit + (1 - delta) * pr$credit)
}

# --- 8. CRRA utility and EXACT expected utility -----------------------------
utility <- function(w, gamma) {
  if (abs(gamma - 1) < 1e-9) log(w) else (w^(1 - gamma)) / (1 - gamma)
}

expected_utility_exact <- function(pf, i, delta, gamma, beta = 1) {
  a_i <- pf$contributions[i]; q_i <- pf$qx[i]
  cc  <- combo_one(pf, i, delta)
  loo <- pool_distribution(pf, setdiff(seq_len(nrow(pf)), i))
  idx_survive <- match(round(loo$value,       8), round(cc$s, 8))
  idx_die     <- match(round(loo$value + a_i, 8), round(cc$s, 8))
  credit_survive <- ifelse(is.na(idx_survive), 0, cc$credit[idx_survive])
  credit_die     <- ifelse(is.na(idx_die),     0, cc$credit[idx_die])
  U_survive <- sum(utility(a_i + credit_survive, gamma) * loo$prob)
  U_die     <- sum(utility(credit_die,           gamma) * loo$prob)
  (1 - q_i) * U_survive + beta * q_i * U_die
}

# --- 8b. Fast helpers for sweeps (heavy parts computed once per member) ------
# precompute_member() runs the expensive convolutions ONCE; eu_fast() then
# evaluates expected utility for any (delta, gamma, beta) by cheap arithmetic.
# eu_fast() is numerically identical to expected_utility_exact().
precompute_member <- function(pf, i) {
  full <- pool_distribution(pf)
  cm   <- cmrs_one(pf, i, full)
  loo  <- pool_distribution(pf, setdiff(seq_len(nrow(pf)), i))
  a_i  <- pf$contributions[i]; q_i <- pf$qx[i]
  w_i  <- (q_i * a_i) / sum(pf$qx * pf$contributions)
  list(a_i = a_i, q_i = q_i,
       cmrs = cm$credit, prop = w_i * cm$s, prob = loo$prob,
       idx_s = match(round(loo$value,       8), round(cm$s, 8)),
       idx_d = match(round(loo$value + a_i, 8), round(cm$s, 8)))
}

eu_fast <- function(P, delta, gamma, beta = 1) {
  credit <- delta * P$cmrs + (1 - delta) * P$prop
  cs <- ifelse(is.na(P$idx_s), 0, credit[P$idx_s])
  cd <- ifelse(is.na(P$idx_d), 0, credit[P$idx_d])
  (1 - P$q_i) * sum(utility(P$a_i + cs, gamma) * P$prob) +
    beta * P$q_i * sum(utility(cd, gamma) * P$prob)
}
```

# The portfolio

```{r portfolio}
pf <- generate_portfolio(n = 100, age_min = 55, age_max = 85,
                         contrib = 50, contrib_sd = 3,
                         prob_male = 0.5, age_dist = 'peaked',
                         age_mean = 70, age_sd = 8, seed = 125)
head(pf)
summary(pf$age)
table(pf$sex)
mean(pf$qx)
```

## Portfolio composition

```{r portfolio-viz, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))

hist(pf$age, breaks = 15, col = "grey80", border = "white",
     main = "Age distribution", xlab = "Age")

barplot(table(pf$sex), col = c("steelblue", "salmon"),
        main = "Sex composition", ylab = "Count")

boxplot(age ~ sex, data = pf, col = c("salmon", "steelblue"),
        main = "Age by sex", xlab = "Sex", ylab = "Age")

hist(pf$qx, breaks = 15, col = "grey80", border = "white",
     main = "Death-probability (qx) distribution", xlab = "qx")

par(mfrow = c(1, 1))
```

# Expected utility vs mixing weight, for several gamma

Each curve is shown **relative to its own maximum** (so every curve has a peak at 0),
because CRRA utility levels differ greatly across gamma. This makes the *shape*
and the location of the optimal `delta` comparable across risk-aversion levels.

```{r multi-gamma-panels, fig.width=10, fig.height=7}
i      <- 1
P      <- precompute_member(pf, i)
deltas <- seq(0, 1, by = 0.02)
gammas <- c(1.5, 3, 5, 8, 12, 20)     # six values -> 2 rows x 3 cols

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))   # 2 rows, 3 columns

for (g in gammas) {
  eu <- sapply(deltas, function(d) eu_fast(P, d, gamma = g, beta = 1))
  d_star <- deltas[which.max(eu)]
  plot(deltas, eu, type = "l", lwd = 2,
       xlab = expression(delta), ylab = "Expected utility",
       main = bquote(gamma == .(g) ~ "," ~ delta*"*" == .(d_star)))
  abline(v = d_star, col = "red", lty = 2)   # mark the optimum
}

par(mfrow = c(1, 1))
```

# Optimal mixing weight as a function of risk aversion

For each `gamma`, the value of `delta` maximizing expected utility. This shows whether the optimal blend of CMRS and proportional shifts
with risk aversion, and whether it stays interior or hits a corner.

```{r optimal-delta, fig.width=8, fig.height=5}
gammas_fine <- seq(1, 12, by = 0.5)
deltas_fine <- seq(0, 1, by = 0.01)

opt_delta <- sapply(gammas_fine, function(g) {
  eu <- sapply(deltas_fine, function(d) eu_fast(P, d, gamma = g, beta = 1))
  deltas_fine[which.max(eu)]
})

plot(gammas_fine, opt_delta, type = "b", pch = 19, lwd = 2,
     ylim = c(0, 1),
     xlab = expression(gamma), ylab = expression(delta * "*"),
     main = paste0("Member ", i, ": optimal mixing weight vs risk aversion"))
abline(h = c(0, 1), lty = 3, col = "grey60")
```

# Effect of the bequest weight

The same optimal-`delta` curve for two bequest weights `beta`, to see how a
weaker bequest motive changes the optimal blend.

```{r bequest, fig.width=8, fig.height=5}
opt_delta_beta <- function(beta) {
  sapply(gammas_fine, function(g) {
    eu <- sapply(deltas_fine, function(d) eu_fast(P, d, gamma = g, beta = beta))
    deltas_fine[which.max(eu)]
  })
}

od1   <- opt_delta_beta(1.0)
od08  <- opt_delta_beta(0.8)

plot(gammas_fine, od1, type = "b", pch = 19, lwd = 2, ylim = c(0, 1),
     xlab = expression(gamma), ylab = expression(delta * "*"),
     main = "Optimal delta vs gamma, by bequest weight")
lines(gammas_fine, od08, type = "b", pch = 17, lwd = 2, col = "red")
legend("bottomleft", legend = c("beta = 1.0", "beta = 0.8"),
       col = c("black", "red"), pch = c(19, 17), lwd = 2, bty = "n")
```
